---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crashcounter

<!-- badges: start -->
[![codecov](https://codecov.io/gh/jwood-iastate/crashcounter/branch/master/graph/badge.svg?token=ofsyBpE669)](https://codecov.io/gh/jwood-iastate/crashcounter)
<!-- badges: end -->

The goal of crashcounter is to provide simple functions for use in preparing crash data for count regression modeling. Users should be able to get crash counts for road segments and intersections (total crashes, crashes that meet specific conditions, etc.) using the function(s) in this package.

## Installation

You can install the development version of crashcounter like so:

``` r
library(remotes)
remotes::install_github("jwood-iastate/crashcounter")
```

## Example: Road Segments using Linear Referencing System (LRS)
Function: `crashCounts.seg()`

This function merges a roads dataset with a crashes dataset based on specified route identifiers and segment boundaries. It aggregates crash data per road segment, allowing for optional user-specified conditions to filter crashes based on various attributes.

```{r example}
library(crashcounter)
set.seed(123) # for reproducibility

roads <- data.frame(
  County = rep(c("A", "B"), each = 4),
  RouteNo = rep(1:2, each = 2, times = 2),
  Region = rep(c("North", "South"), each = 4),
  BeginMeas = c(0, 10, 20, 30, 0, 15, 30, 45),
  EndMeas = c(10, 20, 30, 40, 15, 30, 45, 60),
  NoLanes = sample(2:4, 8, replace = TRUE),
  LaneWidth = runif(8, 3.0, 3.5),
  ShoulderWidth = runif(8, 1.0, 2.5),
  SpeedLimit = sample(c(55, 65, 75), 8, replace = TRUE)
)

crashes <- data.frame(
  County = sample(c("A", "B"), 50, replace = TRUE),
  RouteNo = sample(1:2, 50, replace = TRUE),
  Region = sample(c("North", "South"), 50, replace = TRUE),
  Dist = runif(50, 0, 40),
  Severity = sample(1:5, 50, replace = TRUE),
  Type = sample(c("Collision", "Rollover", "Off-road", "Animal"), 50, replace = TRUE),
  DayofWeek = sample(c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"), 50, replace = TRUE),
  Month = sample(1:12, 50, replace = TRUE),
  Year = sample(2018:2020, 50, replace = TRUE),
  Time = sample(0:23, 50, replace = TRUE),
  NoVehicles = sample(1:4, 50, replace = TRUE),
  NoPeople = sample(1:6, 50, replace = TRUE),
  NoPedestrians = sample(0:2, 50, replace = TRUE)
)

# Run the function
result_basic <- crashCounts.seg(
  roads = roads,
  crashes = crashes,
  road_id_vars = c("County", "RouteNo", "Region"),
  start_mp = "BeginMeas",
  end_mp = "EndMeas",
  crash_mp = "Dist",
  countvarname = "Total_Crashes"
)

# View the result
print(result_basic)

```

For this example, the unique road segments are identified using the combination of:

 - `County`
 - `Region`
 - `RouteNo`
 - `BeginMeas`
 - `EndMeas`
 
In many cases, roadway datasets will have a single variable identifying the route (in this case, the combination of `County`, `Region`, and `RouteNo` are required). The variable `BeginMeas` is the same concept as a starting milepost and `EndMeas` is the same concept as the end milepost for the road segment.

To identify crashes that occur on a given road segment, they have to match `County`, `Region`, and `RouteNo` and have a value for `Dist` that is greater than or equal to `BeginMeas` and less than `EndMeas`. To get specific crash types, conditions can be added.

To illustrate how to add other crash types and use conditions, several additional counts are added below.

Adding Fatal+Injury Crashes - Illustrating the use of a single condition:
```{r}
result_severity <- crashCounts.seg(
  roads = result_basic, # Use the roads dataframe that has the total crashes
  crashes = crashes,
  road_id_vars = c("County", "RouteNo", "Region"),
  start_mp = "BeginMeas",
  end_mp = "EndMeas",
  crash_mp = "Dist",
  conditions = "Severity > 2",
  countvarname = "FatalInjury_Crashes"
)

# View the result
print(result_severity)
```

Next, add more conditions (in this case, nighttime fatal+injury crashes):

```{r}
# Aggregation with multiple conditions: Count crashes with severity greater than 2 and during nighttime
result_multiple <- crashCounts.seg(
  roads = result_severity,
  crashes = crashes,
  road_id_vars = c("County", "RouteNo", "Region"),
  start_mp = "BeginMeas",
  end_mp = "EndMeas",
  crash_mp = "Dist",
  conditions = list(
    expression(Severity > 2),
    expression(Time < 6 | Time > 18)  # Assuming nighttime is before 6 AM or after 6 PM
  ),
  countvarname = "NightFatalInj_Crashes"
)

# View the result
print(result_multiple)
```

## Example: Segments using spatial data
Function: `crashCounts.seg.gdb()`

This function is similar to `crashCounts.seg()` but is designed to work with spatial data stored in a file geodatabase (GDB). The function reads the road and crash data from a GDB and returns the aggregated crash data per road segment.
```{r}
library(sf)
# Sample road segments data with route IDs
road_segments_df <- data.frame(
  segment_id = 1:3,
  route_id = c("Route66", "I-5", "I-10")
)
# Create simple lines for the example
road_segments_df$geometry <- st_sfc(
  st_linestring(rbind(c(-118.25, 34.05), c(-118.24, 34.05))),
  st_linestring(rbind(c(-118.24, 34.05), c(-118.24, 34.06))),
  st_linestring(rbind(c(-118.24, 34.06), c(-118.23, 34.06)))
)
road_segments_sf <- st_sf(road_segments_df, crs = 4326)

# Sample crash data with route IDs
crashes_df <- data.frame(
  crash_id = 1:5,
  crash_latitude = c(34.0505, 34.055, 34.065, 34.051, 34.062),
  crash_longitude = c(-118.245, -118.242, -118.235, -118.243, -118.238),
  Severity = c(1, 3, 2, 5, 4),
  Time = c(2, 15, 20, 5, 23),
  route_id = c("Route66", "I-5", "I-10", "I-5", "Route66")
)

# Run the function
result <- crashCounts.seg.gdb(
  road_segments = road_segments_sf,
  crashes = crashes_df,
  crash_lat = "crash_latitude",
  crash_long = "crash_longitude",
  road_id_vars = c("segment_id"),
  crashRoute = "route_id",
  roadRoute = "route_id",
  dist_feet = 75,
  projection_input = 4326,
  projection_working = 3857
)

# View the result
print(result)
```

# Example: Intersections
Function: `crashCounts.int()`

This function merges a roads dataset with a crashes dataset based on specified intersection identifiers. It aggregates crash data per intersection, allowing for optional user-specified conditions to filter crashes based on various attributes.

```{r}
# Sample data frames
intersections_df <- data.frame(
  intersection_id = 1:3,
  latitude = c(34.0522, 34.0525, 34.0528),
  longitude = c(-118.2437, -118.2440, -118.2443)
)

crashes_df <- data.frame(
  crash_id = 1:5,
  crash_latitude = c(34.0520, 34.0526, 34.0527, 34.0510, 34.0530),
  crash_longitude = c(-118.2430, -118.2441, -118.2442, -118.2445, -118.2446),
  Severity = c(1, 3, 2, 5, 4),
  Time = c(2, 15, 20, 5, 23)
)

# Run the function
result <- crashCounts.int(
  intersections = intersections_df,
  crashes = crashes_df,
  int_lat = "latitude",
  int_long = "longitude",
  crash_lat = "crash_latitude",
  crash_long = "crash_longitude",
  int_id_vars = c("intersection_id"),
  dist_feet = 500,
)

# View the result
print(result)
```
